Architectures for workflow integration in science gateways

### 2016-04-12 paper.tex
Renamed images folder into figures
Wrote section 4.3

@@

\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{images/tight.pdf_tex}
\caption{Tight integration}
\label{archi:tight}
\end{figure}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway. The science gateway controls it
through a specific interaction that might be implemented as a
web-service call (e.g. RESTful or SOAP), as a command-line or as any
other method that separates the workflow engine from the science
gateway. The workflow engine might be invoked either as a black box
that completely masks the infrastructure and workflow activities, or
as a grey box that allows for some interaction with them.  The
workflow engine is responsible for controlling the tasks on the
infrastructure, and performing the required data transfers to execute
them. User data is usually managed through the science gateway,
although it might as well be delivered by the workflow engine directly
to the user.

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that use it~\cite{kacsuk2014science}.
\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{images/service.pdf_tex}
\caption{Service integration}
\label{archi:service}
\end{figure}

@@

This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
integrate the PSOM workflow engine and the FSL toolkit through its
\texttt{fsl\_sub} tool. In particular, the CBRAIN-PSOM integration is
described in~\cite{GLAT-16} and uses an agent computing model (a.k.a pilot
jobs).
\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{images/sub-task.pdf_tex}
\caption{Sub-tasking}
\label{archi:sub-task}
\end{figure}

\subsection{Pool model}
\label{sec:pool}
See Figure~\ref{archi:agent}. Workflows are submitted by the science
gateway to a pool to which agents connect asynchronously to retrieve
and execute workflows (interaction \texttt{f}). Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. This model was implemented in the
SHIWA pool~\cite{ROGE-13}.
\begin{figure}
\centering
\def\svgwidth{0.8\columnwidth}
\input{images/agent.pdf_tex}
\caption{Pool model}
\label{archi:agent}
\end{figure}

@@

\begin{figure}
\centering
\def\svgwidth{0.48\columnwidth}
\input{images/nested-1.pdf_tex}
\def\svgwidth{0.48\columnwidth}
\input{images/nested-2.pdf_tex}
\caption{Nested workflows. Left: abstract model. Right: instantiation with service invocation.}
\label{archi:nested}
\end{figure}

@@

\begin{figure}
\centering
\def\svgwidth{0.48\columnwidth}
\input{images/import-1.pdf_tex}
\def\svgwidth{0.48\columnwidth}
\input{images/import-2.pdf_tex}
\caption{Workflow import. Left: abstract model. Right: instantiation with service invocation.}
\label{archi:import}
\end{figure}

@@

\subsection{Criteria}

@@

\emph{Robustness} relates to the complexity of the architecture and
its ability to provide relevant debugging information about the
workflow executions. It breaks down to 3 metrics:
\begin{itemize}[itemsep=0cm]
\item Components: number of unique components involved in workflow executions.
\item Interactions: number of unique interactions involved in workflow executions. 
\item Debugging: availability of debugging information about workflow activities
  (1: difficult to obtain; 0: easy to access).
\end{itemize}

\emph{Modularity} measures the ability to replace elements in the
architecture, or to integrate new elements. Elements include
workflow engines, workflows and infrastructure. It breaks down in 4
metrics:
\begin{itemize}[itemsep=0cm]
\item New engine type: number of interactions or components to modify
  to integrate a new workflow engine type in the architecture.
\item New engine version: number of interactions or components to
  modify to integrate a new version of a workflow engine in the
  architecture, assuming that another version of the same engine type
  is already available.
\item New infrastructure: number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture.
\item New workflow: number of interactions required to integrate a new
  workflow in the architecture.
\item Meta-workflow: ability to describe meta-workflows from existing
  workflows (1: not available; 0: available).
\end{itemize}

\emph{Scalability} measures the ability of the architecture to support
high workloads. It breaks down into 4 metrics:
\begin{itemize}[itemsep=0cm]
\item New engine instance (0: automated; 1: manual intervention needed).
\item Elastic engines (0: easy to implement; 1: difficult to implement).
\item Distributed engines (0: possible; 1: not possible).
\item Task scheduling (0: difficult; 1: more difficult).
\end{itemize}

@@

\subsection{Evaluation}

\paragraph{Service invocation} Integrating a workflow engine as a
service requires 3 interactions. First, the science gateway has to be
extended to call the engine interface (\texttt{d}), which might
require some adaptation on the workflow engine too, for instance
wrapping in a web service. In addition, the workflow engine has to be
interfaced with the infrastructure, which requires the two
interactions \texttt{b} and \texttt{$c_2$}.

+ Easy to implement in the SG (just call the service)
- Load-balancing between services is difficult
(refer to VIP's experience).
- Need to wrap workflow engines in
services.
- Need to know where to submit each workflow, e.g. when
multiple workflows are involved.
- The science gateway has no
knowledge of the detailed progress of the workflow, in particular
job-level information (statuses, stdout, stderr). If we wanted to do
that, specific feedback channels have to be developed, which requires more work (see VIP).

\paragraph{Sub-task}

@@

\paragraph{Pool}

@@

\begin{table*}
\centering
\begin{tabular}{rcccccc}
\textbf{Metric}                      & \textbf{Tight}
                                     & \textbf{Service}
                                     & \textbf{Sub-task}
                                     & \textbf{Pool}
                                     & \textbf{Nested}
                                     & \textbf{Import} \\
\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
From science gateway                 & \multirow{2}{*}{3}
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99FF99}2  
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}3 \\
From workflow engine                 &
                                     & \cellcolor[HTML]{99FF99}1 
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99FF99}1 \\
\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\

@@

Components                           & \cellcolor[HTML]{99FF99}2

@@

Interactions                         & \cellcolor[HTML]{99FF99}2    

@@

Debugging                            & \cellcolor[HTML]{99AA99}0          

@@

@@

New engine type                      & \cellcolor[HTML]{99AA99}3

@@

New engine version                   & \cellcolor[HTML]{99AA99}1

@@

New infrastructure                   & \cellcolor[HTML]{99FF99}2

@@

New workflow                         & \cellcolor[HTML]{99FF99}1

@@

Meta-workflow                        & \cellcolor[HTML]{99AA99}1

@@

@@

New engine instance                  & \cellcolor[HTML]{99AA99}1

@@

Elastic engines                      & \cellcolor[HTML]{99AA99}2

@@

Distributed engines                  & \cellcolor[HTML]{99AA99}1

@@

Task scheduling                      & \cellcolor[HTML]{99FF99}0

@@

                                     & \cellcolor[HTML]{99BB99}\textbf{18}

@@

>>>>

\todo{Say that the review is done based on our experience with VIP,
  CBRAIN, and to some extent SHIWA.}

@@

\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{figures/tight.pdf_tex}
\caption{Tight integration}
\label{archi:tight}
\end{figure}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction that might be
implemented as a web-service call (e.g. RESTful or SOAP), as a
command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
workflow activities, or as a grey box that allows for some interaction
with them.  The workflow engine is responsible for controlling the
tasks on the infrastructure, and performing the required data
transfers to execute them. User data is usually managed through the
science gateway, although it might as well be delivered by the
workflow engine directly to the user.

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that use it~\cite{kacsuk2014science}.
\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{figures/service.pdf_tex}
\caption{Service integration}
\label{archi:service}
\end{figure}

@@

This model is implemented in CBRAIN~\cite{SHER-14} where it is used to
integrate the PSOM workflow engine and the FSL toolkit through its
\texttt{fsl\_sub} tool. In particular, the CBRAIN-PSOM integration is
described in~\cite{GLAT-16} and uses an agent computing model (a.k.a pilot
jobs).
\begin{figure}
\centering
\def\svgwidth{0.5\columnwidth}
\input{figures/sub-task.pdf_tex}
\caption{Sub-tasking}
\label{archi:sub-task}
\end{figure}

\subsection{Pool model}
\label{sec:pool}
See Figure~\ref{archi:agent}. Workflows are submitted by the science
gateway to a pool to which agents connect asynchronously to retrieve
and execute workflows (interaction \texttt{f}). Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. This model was implemented in the
SHIWA pool~\cite{ROGE-13}.
\begin{figure}
\centering
\def\svgwidth{0.8\columnwidth}
\input{figures/agent.pdf_tex}
\caption{Pool model}
\label{archi:agent}
\end{figure}

@@

\begin{figure}
\centering
\def\svgwidth{0.48\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.48\columnwidth}
\input{figures/nested-2.pdf_tex}
\caption{Nested workflows. Left: abstract model. Right: instantiation with service invocation.}
\label{archi:nested}
\end{figure}

@@

\begin{figure}
\centering
\def\svgwidth{0.48\columnwidth}
\input{figures/import-1.pdf_tex}
\def\svgwidth{0.48\columnwidth}
\input{figures/import-2.pdf_tex}
\caption{Workflow import. Left: abstract model. Right: instantiation with service invocation.}
\label{archi:import}
\end{figure}

@@

\subsection{Evaluation metrics}

@@

\emph{Robustness} relates to the complexity of the architecture and
its ability to provide relevant debugging information about the
workflow executions. It breaks down to 3 metrics:
\begin{itemize}[itemsep=0cm]
\item Components: number of unique components involved in workflow executions.
\item Interactions: number of unique interactions involved in workflow executions. 
\item Debugging: availability of debugging information about workflow
  activities (1: difficult to obtain; 0: easy to access). Fine-grained
  information about workflow activity is required to properly
  troubleshoot workflow executions.
\end{itemize}

\emph{Modularity} measures the ability to replace elements in the
architecture, or to integrate new elements. Elements include
workflow engines, workflows and infrastructure. It breaks down in 4
metrics:
\begin{itemize}[itemsep=0cm]
\item New engine type: number of interactions or components to modify
  to integrate a new workflow engine type in the
  architecture. Modification of an interaction is deemed necessary
  when the parameters involved in this interaction are
  modified. Modification of a component is required when its code
  needs to be modified or recompiled.
\item New engine version: number of interactions or components to
  modify to integrate a new version of a workflow engine in the
  architecture, assuming that another version of the same engine type
  is already available. We assume that different versions of a
  workflow engine share the same interface, i.e. they can be invoked
  using the same software. When this is not the case, the different
  versions have to be considered as different engine types.
\item New infrastructure: number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture.
\item New workflow: number of interactions required to integrate a new
  workflow in the architecture.
\item Meta-workflow: ability to describe meta-workflows from existing
  workflows (1: not available; 0: available).
\end{itemize}

\emph{Scalability} measures the ability of the architecture to support
high workloads. It breaks down into 4 metrics:
\begin{itemize}[itemsep=0cm]
\item New engine instance (0: automated; 1: manual intervention needed).
\item Elastic engines (0: easy to implement; 1: difficult to
  implement; 2: not available).
\item Distributed engines (0: possible; 1: not possible).
\item Task scheduling (0: difficult; 1: more difficult).
\end{itemize}

@@

\paragraph{Integration} Integrating a workflow engine as a service
requires to develop 3 interactions and 1 component. First, a workflow
service has to be developed and interfaced with the infrastructure
through interactions \texttt{b} and \texttt{c$_2$}
(\texttt{I$_2$=3}). In addtion, the science gateway has to be extended
to call the engine interface \texttt{d} (\texttt{I$_1$=1}).

\paragraph{Robustness} Workflow execution requires 3 components: the
science gateway, the workflow service, and the infrastructure
(\texttt{R$_1$=3}). It involves the 4 interactions \texttt{b},
\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
(\texttt{R$_2$=4}). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{R$_3$=0}).

\paragraph{Modularity} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{M$_1$=4}). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{M$_2$=0}). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{M$_3$=3}). New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{M$_4$=1}). Meta-workflow are not supported by default
(\texttt{M$_5$=1}).

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$=1}). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available (\texttt{S$_2$=1}). Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_3$=1}). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_4$=0}).

\subsection{Sub-task}

@@

\subsection{Pool}

@@

\begin{table*}
\centering
\begin{tabular}{rcccccc}
\textbf{Metric}                      & \textbf{Tight}
                                     & \textbf{Service}
                                     & \textbf{Sub-task}
                                     & \textbf{Pool}
                                     & \textbf{Nested}
                                     & \textbf{Import} \\
\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99FF99}2  
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}3 \\
Workflow engine -- \texttt{I$_2$}    &
                                     & \cellcolor[HTML]{99FF99}3 
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99FF99}1 \\
\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99FF99}\textbf{4}
                                     & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\

@@

Components  --   \texttt{R$_1$}      & \cellcolor[HTML]{99FF99}2

@@

Interactions -- \texttt{R$_2$}       & \cellcolor[HTML]{99FF99}3    
                                     & \cellcolor[HTML]{99DD99}4

@@

Debugging --    \texttt{R$_3$}       & \cellcolor[HTML]{99AA99}0          

@@

                                     & \cellcolor[HTML]{99CC99}\textbf{7}

@@

New engine type -- \texttt{M$_1$}    & \cellcolor[HTML]{99AA99}3
                                     & \cellcolor[HTML]{99DD99}4

@@

New engine version -- \texttt{M$_2$} & \cellcolor[HTML]{99AA99}1

@@

New infrastructure -- \texttt{M$_3$} & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99FF99}3

@@

New workflow -- \texttt{M$_4$}       & \cellcolor[HTML]{99FF99}1

@@

Meta-workflow  -- \texttt{M$_5$}     & \cellcolor[HTML]{99AA99}1

@@

                                     & \cellcolor[HTML]{99EE99}\textbf{9}

@@

New engine instance -- \texttt{S$_1$}& \cellcolor[HTML]{99AA99}1

@@

Elastic engines -- \texttt{S$_2$}    & \cellcolor[HTML]{99AA99}2

@@

Distributed engines -- \texttt{S$_3$}& \cellcolor[HTML]{99AA99}1

@@

Task scheduling -- \texttt{S$_4$}    & \cellcolor[HTML]{99FF99}0

@@

                                     & \cellcolor[HTML]{99BB99}\textbf{23}

@@

Comparison between architectures, per criterion (robustness, etc).

This ignores pre-existing work (e.g. a workflow engine is already available as a service). Migration across architectures too.
